Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-15 EBDVF-2024-elsevier/paper.tex
updated EBDVF paper using elsevier template

\documentclass[manuscript,screen]{acmart}

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother

%%
%% This is file `sample-manuscript.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `manuscript')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-manuscript.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

%% PANDOC PREAMBLE BEGINS

@@

\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
%% PANDOC PREAMBLE ENDS

\setlength{\parindent}{10pt}
\setlength{\parskip}{0pt}

\hypersetup{
  pdftitle={Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces},
  pdfauthor={Robert David; Vladimir Alexiev; Petar Ivanov},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={red},
  pdfcreator={LaTeX via pandoc, via quarto}}

%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{}
\acmYear{}
\acmDOI{}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[]{}{}{}
\acmPrice{}
\acmISBN{}

%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Raising the Role of Vocabulary Hubs for Semantic Data
Interoperability in Dataspaces}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

  \author{Robert David}
  \orcid{0000-0002-3244-5341}
            \affiliation{%
                  \institution{The Semantic Web Company}
                                                  \country{Austria}
                      }
        \author{Vladimir Alexiev}
  \orcid{0000-0001-7508-7428}
    \author{Petar Ivanov}
  \orcid{0000-0001-8448-1005}
            \affiliation{%
                  \institution{Ontotext}
                                                  \country{Bulgaria}
                      }

%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}
%%  
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Dataspaces are an important enabler for industrial sharing data (either
commercially licensed or private). Europe is investing heavily into
sectoral dataspaces, federation and orchestration platforms like SIMPL,
Eclipse DSC, GXFS, etc. Still, dataspaces enable shared data access, but
do not solve the data interoperability problem. For that, the consumer
would like to see the data from different providers in a harmonized and
semantically integrated form. The Vocabulary Hub service (part of the
IDSA RAM) provides a repository for ontologies and vocabularies. We
describe an approach of raising the role of the vocabulary hub to also
allow richer metadata description (e.g.~the meaning of every column in a
tabular dataset), and binding semantic descriptions to ingested
datasets, thus providing on-the-fly data semantization and easing data
querying. This is achieved through the integration of two commercial
semantic products (PoolParty and GraphDB), leveraging the partnership
between the Semantic Web Company and Ontotext, and is being developed
within the frame of the Digital Europe project UNDERPIN, with
applications to refinery and wind farm data.    
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{dataspaces, semantic interoperability, semantic
technologies, ontologies, vocabulary hub, oil and gas, renewable
energy, refineries, windfarms}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\setlength{\parskip}{-0.1pt}

@@

The European data economy heavily depends on the availability of data
and the technological foundation to make use of it. Different industries
and communities can benefit mutually by sharing data with each other,
thereby supporting their digitization business goals. Machine-learning
systems heavily rely on high volumes of high-quality training data.
Examples are the well-known mobility dataspace and use cases like energy
communities for running simulations. To cope with the challenge of data
sharing, Dataspace approaches like the IDS RAM were introduced to not
only solve the technical aspects of data sharing, but also to establish
methods for data sovereignty so that dataspace participants can
determine themselves how and when others can make use of their data.

For making use of shared data, interoperability is a crucial factor.
While syntactic data exchange is well covered by clearly defined data
formats, there is still the open challenge of semantic interoperability.
IDS RAM is firmly based on semantic metadata for all aspects and actors
of a dataspace. Furthermore, it can leverage semantic interoperability
approaches through the Vocabulary Hub that stores ontologies and
semantic thesauri for data descriptions and can dereference URIs to
provide semantic details of data assets.

In this paper, we present our proposal for a technological solution of
an IDS Vocabulary Hub built on the integration of the GraphDB semantic
graph database and PoolParty Semantic Suite. We not only cover the core
Vocabulary Hub functionalities and responsibilities in a dataspace, but
also extend the idea towards a Semantic Layer for dataspaces, which
provides even more services to easily and semi-automatically manage
semantic interoperability. Thus we raise the role of the Vocabulary Hub
to not only store semantic assets, but also to bind semantic
descriptions dynamically to incoming data, therefore enabling richer
discovery and semantic data integration between datasets.

@@

TODO

@@

Vocabulary Hub Services: - RDF based: both GraphDB and PoolParty store
and manage data based on the RDF data model. RDF is the basic building
stone and underlying vocabulary definitions. - Scalable vocabulary
storage: cluster implementation for higher throughput and a fault
tolerant architecture. - Knowledge graph management: easy management of
vocabularies using a rich web user interface and integration via Web
APIs. Vocabularies can be created and edited based on standards like
RDFS, OWL and SKOS, and existing vocabularies can be reused and combined
to fit specific dataspace needs. - SPARQL service for semantic lookups:
dereference URIs describing data assets to retrieve a semantic
description of the entity from the graph database.

Semantic Layer Services support both unstructured and structured data
assets: - Unstructured data: semantic annotation services for documents
based on concept tagging. PoolParty's KG-based NLP primarily uses SKOS
taxonomies to identify concepts in documents, but can be extended with
OWL ontology elements (classes, properties) for increased
expressiveness. - Structured data: we support structured data
harmonization methods, including semantic mapping and linking SKOS
concepts - Ontology-Based Data Access (OBDA) is used for virtualization
of relational data, i.e.~accessing relational data as semantic data. It
is well suited for semantic querying of large amounts of data, while
keeping the original relational data in place (NoETL approach).
Ontologies are leveraged to construct the OBDA mapping; they are
governed via the Semantic Layer and can harmonize data assets of various
formats using a precise mapping. Furthermore, it can be used to
establish a fine-grained access control, thereby deciding for individual
data elements if they should be exposed ot not. - RDB to RDF Mapping
Language (R2RML) \citep{souripriyadasR2RMLRDBRDF2012} and RDF Mapping
Language (RML) \citep{RMLIntroduction2024} are intended for data
transformation processes within a dataspace, where the original source
is used as a basis to create an RDF representation (ETL approach). -
GraphDB supports both OBDA and R2RML through its integration of ONTOP
\citep{bagosiOntopFrameworkOntology2014}. The same mappings can be used
for NoETL virtual data access, or for ETL (to materialize the semantic
data). - CSV on the Web (CSVW) \citep{swirrlCSVWStandards2024} is a set
of W3C standards that allow detailed semantic description of tabular
data using a JSON ``manifest'' file. - SKOS concepts are used for
structured data in several ways. Concepts can be directly added to RDF
data via semantic links (properties from an ontology), they can be used
to annotate specific literals of an RDF graph, or they can be used to
associate documents with (specific parts of) an RDF graph. - Inference
services: we provide metadata expansion services based on interlinked
vocabularies in combination with ontologies. This inference approach can
be used with semantic tagging for unstructured data, as well as the
structured data approaches. - Inference tagging is provided by a
PoolParty tagging service in combination with an expansion query for
implementing the inference. The query can be configured to work for
various scenarios. - Vocabulary crosswalks are a data modeling approach
using interlinks to relate: - Different vocabularies from similar or the
same domain that cover different areas or aspects of the domain. By
relating concepts with each other using 1:1, 1:n and m:n relations, we
can automatically expand queries and annotations by traversing the
interlinked graph. - Different versions of the same vocabulary to
automatically translate between those versions. - Metadata inference
based on interlinked vocabularies is similar to crosswalks, but intended
to represent different levels of abstraction, so that we are able to
infer from a more general vocabulary to a more specific vocabulary and
vice versa. For example, we can start with DCAT themes vocabularies and
link to further vocabularies, which are more specific for certain areas
or which go into more detail.

@@

The next steps include: - Bringing it into actual dataspaces in
practice, representing and integrating datasets from different dataspace
participants. - Implementing use cases for richer discoverability,
harmonized querying and support for different content types of
structured and unstructured data. - Explore how ML can benefit regarding
quality in practice when providing consolidated and cleaned data via
dataspaces. - Discuss how we can extend the IDS-RAM with services to
improve the support for semantic interoperability provided by our
solution.

@@

This work is partially supported by the Digital Europe programme project
\href{https://underpinproject.eu/}{UNDERPIN} (grant agreement 101123179)

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography.bib}

%% begin pandoc before-bib
%% end pandoc before-bib
%% begin pandoc biblio
%% end pandoc biblio
%% begin pandoc include-after
%% end pandoc include-after
%% begin pandoc after-body
%% end pandoc after-body

\end{document}
\endinput
%%
%% End of file `sample-manuscript.tex'.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  super,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Workshop on Semantic Interoperability in Data Spaces}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces},
  pdfauthor={Robert David; Petar Ivanov; Vladimir Alexiev},
  pdfkeywords={dataspaces, semantic interoperability, semantic
technologies, ontologies, vocabulary hub, oil and gas, renewable
energy, refineries, windfarms},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Raising the Role of Vocabulary Hubs for Semantic Data
Interoperability in Dataspaces}
\author[1]{Robert David%
%
}
 \ead{robert.david@semantic-web.com} 
\author[2]{Petar Ivanov%
%
}
 \ead{petar.ivanov@ontotext.com} 
\author[2]{Vladimir Alexiev%
%
}
 \ead{vladimir.alexiev@ontotext.com}

\affiliation[1]{organization={The Semantic Web
Company},country={Austria},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Ontotext},country={Bulgaria},countrysep={,},postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
Dataspaces are an important enabler for industrial sharing data (either
commercially licensed or private). Europe is investing heavily into
sectoral dataspaces, federation and orchestration platforms like SIMPL,
Eclipse DSC, GXFS, etc. Still, dataspaces enable shared data access, but
do not solve the data interoperability problem. For that, the consumer
would like to see the data from different providers in a harmonized and
semantically integrated form. The Vocabulary Hub service (part of the
IDSA RAM) provides a repository for ontologies and vocabularies. We
describe an approach of raising the role of the vocabulary hub to also
allow richer metadata description (e.g.~the meaning of every column in a
tabular dataset), and binding semantic descriptions to ingested
datasets, thus providing on-the-fly data semantization and easing data
querying. This is achieved through the integration of two commercial
semantic products (PoolParty and GraphDB), leveraging the partnership
between the Semantic Web Company and Ontotext, and is being developed
within the frame of the Digital Europe project UNDERPIN, with
applications to refinery and wind farm data.
\end{abstract}

\begin{keyword}
    dataspaces \sep semantic interoperability \sep semantic
technologies \sep ontologies \sep vocabulary hub \sep oil and
gas \sep renewable energy \sep refineries \sep 
    windfarms
\end{keyword}
\end{frontmatter}

@@

The European data economy heavily depends on the availability of data
and the technological foundation to make use of it. Different industries
and communities can benefit mutually by sharing data with each other,
thereby supporting their digitization business goals. Machine-learning
systems heavily rely on high volumes of high-quality training data.
Examples are the well-known mobility dataspace and use cases like energy
communities for running simulations. To cope with the challenge of data
sharing, Dataspace approaches like the IDS Reference Architecture Model
(IDS RAM) were introduced to not only solve the technical aspects of
data sharing, but also to establish methods for data sovereignty so that
dataspace participants can determine themselves how and when others can
make use of their data.

For making use of shared data, interoperability is a crucial factor.
While syntactic data exchange is well covered by clearly defined data
formats, there is still the open challenge of semantic interoperability
\citep{woutervandenbergSemanticInteroperabilityData2024}. IDS RAM is
firmly based on semantic metadata for all aspects and actors of a
dataspace. Furthermore, it can leverage semantic interoperability
approaches through the Vocabulary Hub \citep{IDSARAMVocabulary2024} that
stores ontologies and semantic thesauri for data descriptions and can
dereference URIs to provide semantic details of data assets.

In this paper, we present our proposal for a technological solution of
an IDS Vocabulary Hub built on the integration of the GraphDB semantic
graph database and PoolParty Semantic Suite. We not only cover the core
Vocabulary Hub functionalities and responsibilities in a dataspace, but
also extend the idea towards a Semantic Layer for dataspaces, which
provides even more services to easily and semi-automatically manage
semantic interoperability. Thus we raise the role of the Vocabulary Hub
\citep{stornebrinkmichielVocabularyHubSemantic2023} to not only store
semantic assets, but also to bind semantic descriptions dynamically to
incoming data, therefore enabling richer discovery and semantic data
integration between datasets.

@@

Ontotext GraphDB is a highly efficient, scalable and robust graph
database with RDF and SPARQL support. It also offers data virtualization
capabilities which enable direct access using SPARQL to relational
databases, eliminating the need to replicate data. It can perform
semantic inferencing at scale, which enables deriving new semantic facts
from existing facts.

GraphDB handles massive loads, queries, and inferencing in real time.
Reasoning and query evaluation are performed over a persistent storage
layer. Loading, reasoning, and query evaluation are executed extremely
fast even against huge ontologies and knowledge bases. According to the
\href{https://ldbcouncil.org/benchmarks/spb/}{LDBC Semantic Publishing
Benchmark}, it is one of the most scalable OWL repositories currently
available.

@@

Vocabulary Hub Services:

\begin{itemize}
\tightlist
\item
  RDF based: both GraphDB and PoolParty store and manage data based on
  the RDF data model. RDF is the basic building stone and underlying
  vocabulary definitions.
\item
  Scalable vocabulary storage: cluster implementation for higher
  throughput and a fault tolerant architecture.
\item
  Knowledge graph management: easy management of vocabularies using a
  rich web user interface and integration via Web APIs. Vocabularies can
  be created and edited based on standards like RDFS, OWL and SKOS, and
  existing vocabularies can be reused and combined to fit specific
  dataspace needs.
\item
  SPARQL service for semantic lookups: dereference URIs describing data
  assets to retrieve a semantic description of the entity from the graph
  database.
\end{itemize}

Semantic Layer Services support both unstructured and structured data
assets:

\begin{itemize}
\tightlist
\item
  Unstructured data: semantic annotation services for documents based on
  concept tagging. PoolParty's KG-based NLP primarily uses SKOS
  taxonomies to identify concepts in documents, but can be extended with
  OWL ontology elements (classes, properties) for increased
  expressiveness.
\item
  Structured data: we support structured data harmonization methods,
  including semantic mapping and linking SKOS concepts

  \begin{itemize}
  \tightlist
  \item
    Ontology-Based Data Access (OBDA) is used for virtualization of
    relational data, i.e.~accessing relational data as semantic data. It
    is well suited for semantic querying of large amounts of data, while
    keeping the original relational data in place (NoETL approach).
    Ontologies are leveraged to construct the OBDA mapping; they are
    governed via the Semantic Layer and can harmonize data assets of
    various formats using a precise mapping
    \citep{andreselAdaptingOntologybasedData2024}. Furthermore, it can
    be used to establish a fine-grained access control, thereby deciding
    for individual data elements if they should be exposed ot not.
  \item
    RDB to RDF Mapping Language (R2RML)
    \citep{souripriyadasR2RMLRDBRDF2012} and RDF Mapping Language (RML)
    \citep{RMLIntroduction2024} are intended for data transformation
    processes within a dataspace, where the original source is used as a
    basis to create an RDF representation (ETL approach).
  \item
    GraphDB supports both OBDA and R2RML through its integration of
    ONTOP \citep{bagosiOntopFrameworkOntology2014}. The same mappings
    can be used for NoETL virtual data access, or for ETL (to
    materialize the semantic data).
  \item
    CSV on the Web (CSVW) \citep{swirrlCSVWStandards2024} is a set of
    W3C standards that allow detailed semantic description of tabular
    data using a JSON ``manifest'' file.
  \item
    SKOS concepts are used for structured data in several ways. Concepts
    can be directly added to RDF data via semantic links (properties
    from an ontology), they can be used to annotate specific literals of
    an RDF graph, or they can be used to associate documents with
    (specific parts of) an RDF graph.
  \end{itemize}
\item
  Inference services: we provide metadata expansion services based on
  interlinked vocabularies in combination with ontologies. This
  inference approach can be used with semantic tagging for unstructured
  data, as well as the structured data approaches.

  \begin{itemize}
  \tightlist
  \item
    Inference tagging is provided by a PoolParty tagging service in
    combination with an expansion query for implementing the inference.
    The query can be configured to work for various scenarios.
  \end{itemize}
\item
  Vocabulary crosswalks are a data modeling approach using interlinks to
  relate:

  \begin{itemize}
  \tightlist
  \item
    Different vocabularies from similar or the same domain that cover
    different areas or aspects of the domain. By relating concepts with
    each other using 1:1, 1:n and m:n relations, we can automatically
    expand queries and annotations by traversing the interlinked graph.
  \item
    Different versions of the same vocabulary to automatically translate
    between those versions.
  \end{itemize}
\item
  Metadata inference based on interlinked vocabularies is similar to
  crosswalks, but intended to represent different levels of abstraction,
  so that we are able to infer from a more general vocabulary to a more
  specific vocabulary and vice versa. For example, we can start with
  DCAT themes vocabularies and link to further vocabularies, which are
  more specific for certain areas or which go into more detail.
\end{itemize}

@@

The next steps include:

\begin{itemize}
\tightlist
\item
  Bringing it into actual dataspaces in practice, representing and
  integrating datasets from different dataspace participants.
\item
  Implementing use cases for richer discoverability, harmonized querying
  and support for different content types of structured and unstructured
  data.
\item
  Explore how ML can benefit regarding quality in practice when
  providing consolidated and cleaned data via dataspaces.
\item
  Discuss how we can extend the IDS-RAM with services to improve the
  support for semantic interoperability provided by our solution.
\end{itemize}

@@

This work is partially supported by the Digital Europe programme project
\href{https://underpinproject.eu/}{UNDERPIN} (grant agreement
101123179).

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}
